---
title: "Chapter 1, Part 1"
author: "Victoria Hermanson"
format: html
  self-contained: true 
execute:
  echo: true               
  warning: false           
  message: false
editor: visual
---

## Chapter 1, Part 1/2

### Basic summary stats –\> clustering –\> gap statistic

#### Read in files

```{r}
#| label: read in files

# load libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggmap)
library(kableExtra)
library(maps)
library(stringr)
library(circular)
library(viridis)

predator_data <- read.csv("../data/underway_obvs_filterednopick.csv") %>%
  filter(!is.na(Species), Species != "NONE") %>%
  filter(Number != 999)

predator_data$Number <- as.numeric(predator_data$Number)


grid_data <- read.csv("../data/LTERStns_edited.csv")
## merge UNSK with SPSK
predator_data <- predator_data %>%
  mutate(Species = case_when(
    Species == "UNSK" ~ "SPSK",  # replace UNSK with SPSK
    TRUE ~ Species               
  ))


# you have distance calculated, grid lines, and Julian day, hr, min
```

## Species Abundance

```{r}
#| label: Species Abundance

# summarize total sightings per species
species_summary <- predator_data %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(total_sightings = sum(as.numeric(Number), na.rm = TRUE)) %>%
  arrange(desc(total_sightings))

#dev.off()
# total sightings across all species
total_sightings_all <- sum(species_summary$total_sightings)
# number of unique species
num_species <- nrow(species_summary)
# top 5
top_species <- head(species_summary, 5)


unknown_species_summary <- predator_data %>%
  filter(str_starts(Species, "UN")) %>%
  mutate(Number = as.numeric(Number)) %>%
  filter(!is.na(Number)) %>%
  group_by(Species) %>%
  summarise(total_sightings = sum(Number), .groups = "drop") %>%
  arrange(desc(total_sightings))

# total sightings across all species
total_un_sightings_all <- sum(unknown_species_summary$total_sightings)
# number of unique species
num_un_species <- nrow(unknown_species_summary)
# top 5
top_un_species <- head(unknown_species_summary, 5)

ggplot(species_summary, aes(x = reorder(Species, -total_sightings), y = total_sightings)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_y_log10() +  # <-- log-transform y-axis
  labs(title = "Total Sightings per Species (log scale)",
       x = "Species",
       y = "Log10 Total Sightings") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 9))

#ggsave("totalspp_sightings_log_plot.png", width = 10, height = 6, dpi = 300)

```

## Species Frequency

```{r}
#| label: Species Frequency

# species frequency (number of observations per species per year)
frequency_by_year <- predator_data %>%
  group_by(Species, Cruise) %>%
  summarise(frequency = n(), .groups = "drop")

# linear model for each species (only those observed in ≥3 years)
# frequency_trends <- frequency_by_year %>%
#   group_by(Species) %>%
#   filter(n() >= 4) %>%
#   summarise(
#     slope = coef(lm(frequency ~ Cruise))[["Cruise"]],
#     p_value = summary(lm(frequency ~ Cruise))$coefficients[2, 4],
#     .groups = "drop"
#   )

# total number of events spp was seen
total_frequency <- predator_data %>%
  group_by(Species) %>%
  filter(!is.na(Species)) %>%
  summarise(total_frequency = n(), .groups = "drop") %>%
  arrange(desc(total_frequency))

# how many years spp was seen throughout dataset
years_present <- predator_data %>%
  group_by(Species) %>%
  summarise(years_seen = n_distinct(Cruise), .groups = "drop")

# total and number of yrs spp was seen
species_freq_summary <- total_frequency %>%
  left_join(years_present, by = "Species") %>%
  arrange(desc(total_frequency))

```

## Species Richness

```{r}
#| label: Species Richness

richness_by_year <- predator_data %>%
  filter(!is.na(Species), !is.na(Cruise)) %>%    
  group_by(Cruise) %>%
  summarise(species_richness = n_distinct(Species), .groups = "drop")

min_richness <- min(richness_by_year$species_richness)
max_richness <- max(richness_by_year$species_richness)
mean_richness <- mean(richness_by_year$species_richness)
year_min <- richness_by_year$Cruise[which.min(richness_by_year$species_richness)]
year_max <- richness_by_year$Cruise[which.max(richness_by_year$species_richness)]

# simple linear model
model <- lm(species_richness ~ Cruise, data = richness_by_year)
summary(model)

spprichness_lm <- ggplot(richness_by_year, aes(x = Cruise, y = species_richness)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Species Richness by Cruise",
       x = "Cruise",
       y = "Species Richness")

ggsave("species_richness_by_cruise.png", plot = spprichness_lm, width = 8, 
       height = 6, dpi = 300)

```

## Density Calculations & Plots

```{r}
#| label: Density Calculations & Plots

bird_spp_of_interest <- c("SPSK","WISP","ADPE","ANTE","GIPT","BBAL","LMSA","GHAL",
                          "CAPT","SOFU","GEPE","ANPR","BLPT","BBSP","CHPE","KEGU",
                          "BESH","UNSK","ROAL","WCPE","UNPE","DIPT","SNPT","ANPT",
                          "BRSK","ARTE","UNSP","UNAL","EMPE","TBPR","FAPR","MAPE",
                          "WAAL","WSIP","UNPR","BNSW","SOAL","UNTE","TPPR","AMSH",
                          "SGPE","SKSP","GIPR","UNPT","BBPE")

# strip width: 300m for seabirds, 900m for cetaceans
# converted to km
predator_data <- predator_data %>%
  mutate(Strip.Width = ifelse(Species %in% bird_spp_of_interest, 0.3, 0.9))

# distance is in km (from haversine function)

# calc the area covered for each event
# distance is in km
predator_data <- predator_data %>%
  mutate(Area.Covered = Distance * Strip.Width)

# to account for any distances that were 0 perchance
# predator_data$Number <- as.numeric(predator_data$Number)
# predator_data <- predator_data %>%
#   mutate(Density = ifelse(Area.Covered > 0, Number / Area.Covered, NA))

predator_data <- predator_data %>%
  mutate(
    Number = as.numeric(Number),
    Area.Covered = as.numeric(Area.Covered),
    Density = ifelse(!is.na(Area.Covered) & Area.Covered > 0,
                     Number / Area.Covered,
                     NA_real_)
  )

```

## Long-term mean and variance calcs/plots

```{r}
#| label: Long-term mean and variance calcs/plots

longterm_spp_density <- predator_data %>%
  group_by(Species) %>%
  summarise(
    Mean_Density = mean(Density, na.rm = TRUE),
    SD_Density = sd(Density, na.rm = TRUE),
    .groups = 'drop'
  )

longterm_spp_density <- longterm_spp_density %>%
  mutate(Species = reorder(Species, Mean_Density))

# dot and error bars
ltdens_dotanderror <- ggplot(longterm_spp_density, aes(x = reorder(Species, Mean_Density), y = Mean_Density)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_errorbar(aes(ymin = Mean_Density - SD_Density,
                    ymax = Mean_Density + SD_Density),
                width = 0.2, color = "darkgreen") +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "Species",
    y = "Long-term Mean Density"
  )

# ggsave("longterm_density_dotanderror.png",
#        plot = ltdens_dotanderror,
#        width = 8, height = 6, units = "in", dpi = 300)

# box plots
ltdens_boxplot <- ggplot(predator_data, aes(x = Species, y = Density)) +
  geom_violin(fill = "lightgreen", alpha = 0.5) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  theme_classic(base_size = 14) +
  coord_flip() +
  labs(
    x = "Species",
    y = "Density (across years)"
  )
print(ltdens_boxplot)

# ggsave("longterm_density_boxplot.png",
#        plot = ltdens_boxplot,
#        width = 8, height = 6, units = "in", dpi = 300)

ggplot(longterm_spp_density, aes(x = Species, y = Mean_Density)) +
  geom_col(fill = "lightgreen") +
  geom_errorbar(aes(ymin = Mean_Density - SD_Density, ymax = Mean_Density + SD_Density),
                width = 0.2, color = "darkgreen") +
  theme_classic() +
  labs(
    title = "Long-Term Mean Density per Species",
    x = "Species",
    y = "Mean Density"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ggsave("longterm_density_plot.png", 
#        plot = last_plot(), 
#        width = 10, height = 6, units = "in", dpi = 300)

```

## Annual mean and variance calcs/plots

```{r}
#| label: Annual mean and variance calcs/plots

annual_spp_density <- predator_data %>%
  group_by(Cruise, Species) %>%
  summarise(
    Mean_Density = mean(Density, na.rm = TRUE),
    SD_Density = sd(Density, na.rm = TRUE),
    .groups = 'drop'
  )

print(annual_spp_density)

# plotting
ggplot(annual_spp_density, aes(x = Cruise, y = Mean_Density, group = Species)) +
  geom_ribbon(aes(ymin = Mean_Density - SD_Density,
                  ymax = Mean_Density + SD_Density,
                  fill = Species), alpha = 0.2, colour = NA) +
  geom_line(aes(color = Species), size = 0.5) +
  facet_wrap(~ Species, scales = "free_y") +
  theme_minimal(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),   
    plot.background = element_rect(fill = "white", color = NA),
    axis.text.y = element_text(size = 8),     
  ) +
  labs(
    title = "Annual Mean Density per Species",
    x = "Cruise",
    y = "Mean Density"
  ) +
  theme(
    legend.position = "none",  # each facet only has one species, so no need for a legend
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ggsave("annual_density_plot.png",
#        plot = last_plot(),
#        width = 10, height = 6, units = "in", dpi = 300)

```

## Separate Station.Start

### Wanted to merge predator data with grid data by GridLine & GridStation

### This would assign each observation to a region

```{r}
#| label: Separate Station.Start

# ensure Station.Start is character
predator_data <- predator_data %>%
  mutate(Station.Start = as.character(Station.Start))

# separate gridline and gridstation
predator_data <- predator_data %>%
  mutate(
    # Extract GridLine (before first dot)
    GridLine = as.numeric(str_extract(Station.Start, "^[^\\.]+")),
    # Extract the part after the first dot
    station_part = str_replace(Station.Start, "^[^\\.]+\\.", "")
  ) %>%
  mutate(
    GridStation_raw = case_when(
      # e.g. "14.5"
      str_detect(station_part, "\\.") ~ as.numeric(station_part),
      
      # e.g. "-020"
      str_detect(station_part, "^\\-\\d+$") ~ as.numeric(station_part),
      
      # three or more digits -> direct numeric (e.g. "080" -> 80)
      str_detect(station_part, "^\\d{3,}$") ~ as.numeric(station_part),
      
      # two digits -> multiply by 10 (e.g. "04" -> 40, "14" -> 140)
      str_detect(station_part, "^\\d{2}$") ~ as.numeric(station_part) * 10,
      
      # one digit -> multiply by 100 (e.g. "2" -> 200)
      str_detect(station_part, "^\\d{1}$") ~ as.numeric(station_part) * 100,
      
      TRUE ~ NA_real_
    )
  ) %>%
  mutate(
    # round to nearest 20 (to match bin size)
    GridStation = if_else(!is.na(GridStation_raw),
                          round(GridStation_raw / 20) * 20,
                          NA_real_)
  ) %>%
  select(-station_part, -GridStation_raw)

```

## Separate by region

```{r}

#| label: Separate by region

predator_data <- predator_data %>%
  mutate(
    GridLine = round(GridLine / 100) * 100   # round to nearest 100
  )

# predator_data <- predator_data %>%
#   mutate(
#     GridStation = round(GridStationOrig / 20) * 20,
#     GridLine = round(GridLineOrig / 100) * 100
#     )

predator_data <- predator_data %>%
  mutate(
    GridLine = as.numeric(GridLine),
    GridStation = as.numeric(GridStation)
  )

# merge predator data with station data
seabird_df <- merge(predator_data, grid_data, by = c("GridLine", "GridStation"))

```

## Climatology maps

#### Climatology maps separated by species, with individual plots [here](https://docs.google.com/document/d/1gSkA4JbkxIu9SqdTmdP9s0hqliipCSjqj2N0GFEtHCM/edit?tab=t.9luj0eq0bqta).

```{r}
#| label: Climatology maps

# top 10 species by total density
top10_species <- seabird_df %>%
  group_by(Species) %>%
  summarize(total_density = sum(Density, na.rm = TRUE)) %>%
  arrange(desc(total_density)) %>%
  slice_head(n = 10) %>%
  pull(Species)
# annual mean predator densities per region

# subset data to only include top 10 spp
seabird_top10 <- seabird_df %>%
  filter(Species %in% top10_species)


# climatology maps per spp
climatology <- seabird_df %>%
  group_by(Species, GridLine, GridStation, Longitude.Start, Latitude.Start) %>%
  summarize(mean_density = mean(Density, na.rm = TRUE))


ggplot(climatology, aes(x = Longitude.Start, y = Latitude.Start, color = mean_density)) +
  geom_point(size = 3) +
  scale_color_viridis_c(option = "C", trans = "log10", name = "Mean density (log10)") +
  coord_fixed() +
  facet_wrap(~Species) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))


# # separate plots for each spp
# species_list <- unique(climatology$Species)
# 
# # loop through each species and save 
# for (sp in species_list) {
#   spp_data <- filter(climatology, Species == sp)
#   
# # plot
# separate_clim_eachspp <- ggplot(spp_data, aes(x = Longitude.Start, y = Latitude.Start, color = mean_density)) +
#   geom_point(size = 3, alpha = 0.9) +
#   scale_color_viridis_c(option = "C", trans = "log10", name = "Mean density (log10)") +
#   coord_fixed() +
#   theme_classic(base_size = 14) +
#   theme(
#     panel.grid = element_blank(),
#     panel.background = element_rect(fill = "white", color = NA),
#     strip.text = element_text(face = "bold"),
#     legend.position = "bottom"
#   ) +
#   labs(
#     title = paste("Climatology Map:", sp),
#     x = "Longitude",
#     y = "Latitude"
#   )
# ggsave(
#   filename = paste0("climatology_", gsub(" ", "_", sp), ".png"),
#   plot = separate_clim_eachspp,
#   width = 8,
#   height = 6,
#   dpi = 300
# )
# }



# climatology maps for the top ten spp
climatology_top10 <- seabird_df %>%
  filter(Species %in% top10_species) %>%
  group_by(Species, GridLine, GridStation, Longitude.Start, Latitude.Start) %>%
  summarize(mean_density = mean(Density, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Longitude.Start = as.numeric(Longitude.Start),
    Latitude.Start = as.numeric(Latitude.Start),
    Species = trimws(Species)
  )

topten <- ggplot(filter(climatology_top10, mean_density > 0),
       aes(x = Longitude.Start, y = Latitude.Start, color = mean_density)) +
  geom_point(size = 2.5, alpha = 0.9) +
  scale_color_viridis_c(
    option = "C",
    trans = "log10",
    name = "Mean density (log10)"
  ) +
  facet_wrap(~Species, ncol = 5) +
  coord_quickmap() +
  theme_minimal(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),   # panel area
    plot.background = element_rect(fill = "white", color = NA),    # entire figure
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "white", color = "gray70"),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm")
  ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Predator Climatology Maps (1993–2023)",
    subtitle = "Top 10 species by mean density"
  )

ggsave(
  filename = "seabird_climatology_top10.png", 
  plot = topten,                                 
  width = 16, height = 10,                   
  dpi = 300                                 
)

## single out specific species

target_species <- "ADPE"

climatology_one <- seabird_df %>%
  filter(Species == target_species) %>%
  group_by(GridLine, GridStation, Longitude.Start, Latitude.Start) %>%
  summarize(mean_density = mean(Density, na.rm = TRUE), .groups = "drop")

ggplot(climatology_one, aes(x = Longitude.Start, y = Latitude.Start, color = mean_density)) +
  geom_point(size = 3) +
  scale_color_viridis_c(option = "C", trans = "log10", name = "Mean density (log10)") +
  coord_fixed() +
  labs(title = target_species) +
  theme_minimal()



```

## Linear regression for bins

```{r}
#| label: Linear regression for bins

# first coefficient is intercept
# second coefficient is slope - rate of change per year (trend)
linreg <- seabird_df %>%
  mutate(bin = paste(Offshore, NorthSouthFar, sep = "_")) %>%
  group_by(bin) %>%
  summarise(
# the slope — how fast density changes per year
    slope = if (n_distinct(Cruise) > 1) coef(lm(Density ~ Cruise))[2] else NA_real_,
# the p-value testing if slope differs from zero
    p_value = if (n_distinct(Cruise) > 1) summary(lm(Density ~ Cruise))$coefficients[2, 4] else NA_real_,
# how much of the density variation the line explains (trend strength)
    r_squared = if (n_distinct(Cruise) > 1) summary(lm(Density ~ Cruise))$r.squared else NA_real_
  )

print(linreg)

linreg_top10 <- seabird_df %>%
  mutate(bin = paste(Offshore, NorthSouthFar, sep = "_")) %>%
  group_by(bin) %>%
  summarise(
    # the slope — how fast density changes per year
    slope = if (n_distinct(Cruise) > 1) coef(lm(Density ~ Cruise))[2] else NA_real_,
    # the p-value testing if slope differs from zero
    p_value = if (n_distinct(Cruise) > 1) summary(lm(Density ~ Cruise))$coefficients[2, 4] else NA_real_,
    # how much of the density variation the line explains (trend strength)
    r_squared = if (n_distinct(Cruise) > 1) summary(lm(Density ~ Cruise))$r.squared else NA_real_
  )

print(linreg_top10)

# p-value indicates how statistically significant slope is (< 0.05)

```

## Correlation & clustering

### Notes on top ten species clustering [here](https://docs.google.com/document/d/1gSkA4JbkxIu9SqdTmdP9s0hqliipCSjqj2N0GFEtHCM/edit?tab=t.0).

```{r}
#| label: Correlation & clustering

# mean density per species per bin
species_bin <- seabird_df %>%
  group_by(Species, Offshore, NorthSouthFar) %>%
  summarize(mean_density = mean(Density, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(bin = paste(Offshore, NorthSouthFar, sep = "_")) %>%
  select(Species, bin, mean_density) %>%
  pivot_wider(names_from = bin, values_from = mean_density, values_fill = 0)

# create species-by-bin matrix
species_matrix <- as.data.frame(species_bin)
rownames(species_matrix) <- species_matrix$Species
species_matrix <- species_matrix[, -1]

# find Pearson correlation coefficient (for distance)
# transposes the matrix so that species become columns (since cor() calcs correlations between columns)
# (1-cor) converts correlations into distances (high correlation = small distance)
# as.dost converts matrix into a formal “distance” object for hierarchical clustering
dist_matrix <- as.dist(1 - cor(t(species_matrix), method = "pearson"))
cor_matrix <- cor(t(species_matrix), method = "pearson")
view(cor_matrix)

# cluster
# clusters species based on how similar their density patterns are
# 1 - r = distance
# "average" defines how clusters merge based on average pairwise distance
hc <- hclust(dist_matrix, method = "average")

# plot dendrogram
png("dendrogram.png", width = 800, height =300)
par(mar = c(5, 10, 4, 2))
plot(hc, main = "Species Clustering Dendrogram")
dev.off()

# simple heatmap w/dendrogram 
# heatmap(
#   cor_matrix,
#   Rowv = as.dendrogram(hc),   
#   Colv = as.dendrogram(hc),
#   col = colorRampPalette(c("blue", "white", "red"))(100),
#   scale = "none",
#   margins = c(8, 8),
#   main = "Species Correlation Heatmap"
# )


# blue: strong negative correlation (close to -1)
# white: no correlation (around 0)
# red: strong positive correlation (close to +1)
heatmap(
  cor_matrix,
  Rowv = NA,
  Colv = NA,
  col = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  margins = c(8, 8),
  main = "Species Correlation Heatmap"
)

dev.copy(png, "heatmap.png", width = 800, height = 600)
dev.off()

```

## Gap statistic to find optimal number of clusters

```{r}
#| label: Gap statistic to find optimal number of clusters

# k-means for range of potential clusters
# wss = within-cluster sum of squared errors (inertia?)

# number of clusters to test
k_values <- 1:10
wss_values <- numeric(length(k_values))  # empty vector to keep k_avlues

# loop over k = 1 to 10
for (i in seq_along(k_values)) {
  k <- k_values[i]
# cut tree into k clusters
  groups <- cutree(hc, k = k)
# get list of unique clusters
  cluster_ids <- unique(groups)
# begin WSS for this k
  total_wss <- 0
# loop over each cluster
  for (g in cluster_ids) {
    members <- which(groups == g)                     # indices of members
    distance <- as.matrix(dist_matrix)[members, members]     # distances within cluster
    total_wss <- total_wss + sum(distance^2) / (2 * length(members))
  }
  
  # keep result
  wss_values[i] <- total_wss
}

# plot elbow curve
# the point in between where there is a steep slope is number to use
plot(k_values, wss_values, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Within-cluster sum of squares",
     main = "Elbow Method to Estimate Optimal Number of Clusters")

png("elbow_plot.png", width = 800, height = 600)
plot(k_values, wss_values, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Within-cluster sum of squares",
     main = "Elbow Method to Estimate Optimal Number of Clusters")
dev.off()



```

![]()
